---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
str(activity <- read.csv("activity.csv"))
```

### Convert to an R date field

```{r}
activity$r.date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

### Make a histogram of the total number of steps taken each day
```{r Total_Steps_by_Day}
total.daily.steps <- tapply(activity$steps, activity$r.date, sum, na.rm = TRUE)

hist(total.daily.steps, xlab = "Day", ylab = "Count", main = "Total Steps by Day")
```

### Calculate and report the mean and median total number of steps taken per day
```{r}
mean.steps <- mean(total.daily.steps)
median.steps <- median(total.daily.steps)
```

The mean steps per day is `r mean.steps`.
The median steps per day is `r median.steps`.

## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r Avg_Steps_per_Interval}
interval.steps.mean <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)

plot(interval.steps.mean, type = "l", main = "Avg Steps per Interval", xlab = "Interval Number", ylab = "Avg Steps")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max.avg.steps <- names(which.max(interval.steps.mean))
```

The interval with the highest average number of steps is `r max.avg.steps`.

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
activity.na <- sum(is.na(activity))
```

The total number of rows with NA values is `r activity.na`.

### Impute NA values using the global mean number of steps
```{r}
activity.impute <- activity
require(Hmisc)
activity.impute$steps <- impute(activity.impute$steps, mean)
```

### Repeat earlier analysis using imputed values
```{r Total_Steps_by_Day_Imputed}
total.daily.steps.impute <- tapply(activity.impute$steps, activity.impute$date, sum)

hist(total.daily.steps.impute, xlab = "Day", ylab = "Count", main = "Total Steps by Day (with Inputed)")

mean.steps.impute <- mean(total.daily.steps.impute)
median.steps.impute <- median(total.daily.steps.impute)
```

The mean steps per day after imputation is `r mean.steps.impute`.
The median steps per day after imputation is `r median.steps.impute`.

The imputation appears to have caused summary statistics to be higher.

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activity.impute$date.type <- as.factor(ifelse(weekdays(activity.impute$r.date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))
str(activity.impute)
is.factor(activity.impute$date.type)
```

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r Weekday_vs_Weekend_Trend}
date.type.means <- aggregate(steps ~ date.type + interval, data = activity.impute, mean)

require(lattice)
xyplot(steps ~ interval | date.type, data = date.type.means, layout = c(1,2), type = "l", main = "Comparison of Avg # of Steps by Date Type", xlab = "Interval Number", ylab = "Avg # of Steps")
```
